---
title: "Math261A_Project1"
subtitle: "Subtitle"
author: "Isis Martinez, SJSU ID: "
thanks: "Project repository available at"
date: today
date-format: long
abstract: "This project investiga."
output: 
  pdf_document:
    latex_engine: xelatex
bibliography: references.bib
---

IMPORTANT NOTES
citation: [@palmer_penguins] --> references.bib file
number of rows: `r nrow(adelie)` --> refer to results of analysis directly in text
The estimated slope parameter is $b_1=$ `r round(coefficients(lm_fit)[2],3)` --> refer to slope parameters directly in text`


```{r} 
#can include r and python
#| include: false
#| warning: false
#| message: false 
# load libraries
library(tidyverse)
library(ggpubr)
library(patchwork)


```

```{r, include=FALSE, warning=FALSE, message=FALSE}
#can include r and python
library(tidytuesdayR)
library(dplyr)
library(ggplot2)

```

```{#Predictor (independent variable): unemployment (unemp)}
#Outcome (dependent variable): cardiovascular disease prevalence (cvd)

#other options
#
```
1. Abstract/Introduction
2. Background and data
3. Methods
4. Results
5. Discussion




# Abstract
# Introduction

What is my big question?
You should care about the answer to this question because...
We expect that...
@sec-data introduces the data used in this analysis... @sec-methods describes the model... 


# Data {#sec-load_data}
```{r}
library(tidytuesdayR); library(dplyr); library(ggplot2);library(broom)
library(scales); library(patchwork); library(ggrepel); library(qqplotr)

raw_data <- read.csv("C:/Users/terra/Documents/Schoolwork/MATH261A_Fa25/project1/paper/CalEnviroScreen_4.0_raw_data.csv") # load data

#coerce to numeric & drop NAs
to_analyze_df_3vars <- raw_data %>%
  transmute(
    pov = as.numeric(pov),
    edu = as.numeric(edu),
    unemp = as.numeric(unemp)
  ) %>%
  filter(!is.na(pov), !is.na(edu), !is.na(unemp))

to_analyze_df <- raw_data %>%
  transmute(
    pov   = as.numeric(pov),
    edu   = as.numeric(edu)
    ) %>%
  filter(!is.na(pov), !is.na(edu))
```


```{r fig-education-scatter}
# Model
set.seed(123) 
lm_fit <- lm(pov ~ edu, data=to_analyze_df_3vars) #predictor (x): edu, outcome (y): pov
coefs <- coef(lm_fit)
intercept <- coef(lm_fit)[1]
slope <- coef(lm_fit)[2] 
r2 <- summary(lm_fit)$r.squared

gl <- glance(lm_fit)
n  <- stats::nobs(lm_fit)

eq_text <- sprintf(" regression line: y = %.4f + %.4f x", intercept, slope, r2) # create equation text

#scatter plot with regression
ggplot(to_analyze_df_3vars, aes(x = edu, y = pov, color=unemp)) + #plot
  geom_point(alpha = 0.7, size = .8) +
  scale_color_viridis_c(name = "Unemployment (%)", option = "plasma")+
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  annotate("label", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.1,
           label = eq_text, label.size = .2, alpha = .5) +
  labs(
    title    = "Educations vs. Poverty",
    subtitle = "Simple regression scatter plot of percentage of , colored by unemployment",
    x        = "Adults without a high school diploma (%)",
    y        = "Population under poverty line (%)",
    caption  = "Source: OEHHA CalEnviroScreen 4.0"
  ) +
  theme_pubr()

print(eq_text)
summary(lm_fit)
```

```{r fig-adelie-scatter}
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Scatter plot of flipper length, in mm (x-axis) and body mass in grams (y-axis) with fitted linear regression model."
data(penguins)
adelie <- penguins |>
  filter(species == "Adelie" & sex == "female")
ggplot(adelie, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  # helpful function for adding linear regression line below
  geom_smooth(method = "lm", se  = FALSE) +
  # helpful function from ggpubr package for labeling line below
  stat_regline_equation(label.x = 197, label.y = 3400) +
  theme_pubr() +
  ylab("Body mass (grams)") +
  xlab("Flipper length (mm)") +
  ggtitle("Comparison of flipper length and body mass\nfor female Adelie penguins")
```
  1) describe the observational units of the dataset; 
  2) introduce and explain all relevant variables; 
  3) describe any relevant details about how the data were collected or transformed; 
  4) describe alternative/additional sources of data;
  5) highlight any critical characteristics of the data using visualizations or summary tables. A good rule of thumb: for every variable that is relevant for your analysis, describe how it is calculated, whether there are any limitations related to missingness or measurement error, and provide summary visualizations/tables if
necessary.

# Data {#sec-methods}
Methods. The paper should present any model or statistical methods clearly using
mathematical notation and plain English explanations. Every symbol and component
of the model should be introduced and described. Your target reader should
be a peer in this class. This section must clearly describe any relevant assumptions
and whether they are met. Describe possible limitations of your analysis. This section
should describe the software used for implementation. This section should
also include any necessary steps taken for model selection and validation.

#Results. 
The results section should describe the fitted model and its implications.
This section should include relevant tables, graphs, and quantitative results, as well
as plain English explanations of all of these components. Clearly explain what the
model results show and connect your results to the central research question(s) and
their broader context.
g. References. The text should include in-text citations as well as a reference list for
all sources, including any data, software, and relevant literature. You are advised
to use BiBTeX for compling your references. All citations should be properly formatted
according to an appropriate style guide.



```{r}
head(raw_data)
```
MORE CODE:
```{r}
library(dplyr); library(ggplot2); library(lmtest); library(sandwich)
library(splines); library(broom); library(nlme); library(tidyr)

# --- data
raw <- read.csv("CalEnviroScreen_4.0_Results.csv")
dat <- raw %>% transmute(pov = as.numeric(pov), edu = as.numeric(edu)) %>% drop_na()

# base linear model
m_lin <- lm(pov ~ edu, data = dat)
sum_lin <- summary(m_lin)

# diagnostics: heterosk., nonlinearity
bp <- bptest(m_lin)                                    # heteroskedasticity
m_ns <- lm(pov ~ ns(edu, df = 3), data = dat)          # spline check (nonlinearity)
anova_lin_vs_spline <- anova(m_lin, m_ns)              # p > .05 -> linearity is fine
rst <- resettest(m_lin, power = 2:3, type = "fitted")  # general misspec test

# robust HC3 inference (makes model simple & interpretable)
Vhc3 <- vcovHC(m_lin, type = "HC3")
rob  <- coeftest(m_lin, vcov = Vhc3)
crit <- qt(0.975, df = m_lin$df.residual)
rob_ci <- cbind(Estimate = coef(m_lin),
                `Robust SE (HC3)` = sqrt(diag(Vhc3)),
                `Lower 95%` = coef(m_lin) - crit*sqrt(diag(Vhc3)),
                `Upper 95%` = coef(m_lin) + crit*sqrt(diag(Vhc3)))

print(sum_lin); print(bp); print(anova_lin_vs_spline); print(rst)
print(rob); print(round(rob_ci, 4))

# options 1: transform Y only (variance stabilization, keep linear in x)
m_sqrtY <- lm(sqrt(pov) ~ edu, data = dat)
bp_sqrt <- bptest(m_sqrtY)     # if this is still tiny, robust SE is still fine
summary(m_sqrtY); print(bp_sqrt)

# obtion 2: quadratic term (1 predictor, slight curvature allowed)
m_quad <- lm(pov ~ edu + I(edu^2), data = dat)
anova_lin_vs_quad <- anova(m_lin, m_quad)  # if p < .05, tiny curvature helps
summary(m_quad); print(anova_lin_vs_quad)

# Option 3: simple WLS (variance grows with mean -> weight by 1/fit^2 as a rough fix)
fit0 <- fitted(m_lin)
w    <- 1 / pmax(fit0^2, quantile(fit0^2, 0.01))  # avoid extreme weights
w    <- w / mean(w)
m_wls <- lm(pov ~ edu, data = dat, weights = w)
summary(m_wls); bptest(m_wls)



# Plots (raw model)
eq <- sprintf("y = %.3f + %.3f x (R^2 = %.3f)", coef(m_lin)[1], coef(m_lin)[2], sum_lin$r.squared)
p1 <- ggplot(dat, aes(edu, pov)) +
  geom_point(alpha=.5, size=.6) +
  geom_smooth(method="lm", se=FALSE) +
  annotate("label", x=-Inf, y=Inf, hjust=-0.05, vjust=1.1, label=eq) +
  labs(x="% adults without HS diploma (edu)", y="Poverty (%)",
       title="Poverty vs Education: raw linear fit") +
  theme_minimal()

resid_df <- data.frame(fit=fitted(m_lin), res=resid(m_lin))
p2 <- ggplot(resid_df, aes(fit, res)) +
  geom_point(alpha=.5, size=.6) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_smooth(method="loess", se=FALSE, color="orangered") +
  labs(title="Residuals vs Fitted (raw model)")

p3 <- ggplot(resid_df, aes(sample=res)) +
  stat_qq(size=.6) + stat_qq_line() +
  labs(title="Q–Q Plot (raw model)")

print(p1); print(p2); print(p3)

```

## Including Plots

```{r pressure, echo=FALSE}
plot(pressure)
```

EXTRA CODE:
```{r}
library(dplyr); library(lmtest); library(sandwich); library(broom)

dat <- raw %>%
  transmute(pov = as.numeric(pov), edu = as.numeric(edu)) %>%
  tidyr::drop_na()

m_raw <- lm(pov ~ edu, data = dat)

# OLS summary
summary(m_raw)

# HC3 robust inference
rob_vcov <- vcovHC(m_raw, type = "HC3")
rob_test <- coeftest(m_raw, vcov = rob_vcov)
rob_test

# Robust 95% CIs
crit <- qt(0.975, df = m_raw$df.residual)
rob_se <- sqrt(diag(rob_vcov))
rob_ci <- cbind(
  Estimate = coef(m_raw),
  `Robust SE (HC3)` = rob_se,
  `Lower 95%` = coef(m_raw) - crit*rob_se,
  `Upper 95%` = coef(m_raw) + crit*rob_se
)
round(rob_ci, 4)

m_sqrtY <- lm(sqrt(pov) ~ edu, data = dat)
summary(m_sqrtY)
bptest(m_sqrtY)   # check heteroskedasticity again


library(ggplot2)
eq <- sprintf("y = %.3f + %.3f x (R^2 = %.3f)",
              coef(m_raw)[1], coef(m_raw)[2], summary(m_raw)$r.squared)

ggplot(dat, aes(edu, pov)) +
  geom_point(alpha = .5, size = .6) +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("label", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.1, label = eq) +
  labs(x = "% adults without HS diploma (edu)",
       y = "Poverty (%) (pov)",
       title = "Poverty vs Education (California census tracts)") +
  theme_minimal()


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

Add for references:
Info about data: https://oehha.ca.gov/sites/default/files/media/downloads/calenviroscreen/report/calenviroscreen40reportf2021.pdf


Asthma (rate per 10,000) -->raw, percentile
Cardiovascular Disease (heart attacks per 10,000) -->raw, percentile
Low Birth Weight (percent) -->raw, percentile

